Determinism, Complexity, and Predictability in Computer Performance

### 2014-02-19 pecomplexity.tex
Split Sections to help with conflicts

\section{Introduction}\label{sec:intro}

BRAINSTORMING 
\begin{itemize}
\item Complexity need not be hard to predict (can point at the simple predictions paper) [[move to introduction]]
\item random walk for example is best predicted by guess what just happend[[move to introduction]]
\item The kind of complexity present matters, i.e., that is whether the complexity is structured or not.[[use here and mention in intro]]

\item \col brings about the point nicely that some prediction strategies cannot utilize the processes internal information transfer method. That is a nonlinear internal information transfer system cannot be predicted effectively with a linear strategy. This gives a practitioner leverage on when to give up and when to keep working. [[use in this section as bridge to next section]]

\end{itemize}

%\begin{it}
%Paragraph on computer performance, including citations to Todd paper
%and summary of the results that indicate that they're deterministic
%nonlinear dynamical systems.  Given that, we should be able to
%predict.  What benefits would accrue if we could do so: power mgmt,
%end world hunger [[this is my primary goal everyday :)]], etc.
%\end{it}
Things to add to introduction
\begin{enumerate}
\item Different kinds of complexity exist in time series and this makes choosing prediction models difficult 
\subitem NOTE: RW and chaos are both complex. One is predictable and one is not.

\item Make an argument that Computer Performance is a great testing ground as it omits signals that completely cover the spectrum of complexity \col \dots \gcc

\item When deterministic structure even complex structure exists that structure can be utilized for prediction. 
\item For noisy real-valued time series distinguishing randomness (WN,RW) complexity from structured nonlinear / chaotic /high period / high dimensional etc complexity is (until now) very hard. 
\subitem for this provide predictions of \gcc and \col side by side and discuss "How can we tell if we did a bad job because the method is inadequate vs the signal being too complex. Lead this into is it possible to tell if there exists structure in a time series to know if we should find a better model or not. Maybe even having 4 predictions. top being ARIMA of the above signals and bottom being LMA of the above signals. Show that one improved and one did not. Is it that we used the wrong method to predict or is it that we simply can't predict the signal better than a random walk due to high levels of internal signal complexity.

\item Introduce the two main contributions of the paper which are outlined at the begining of the results section

\end{enumerate}

Computers are among the most complex engineered artifacts in current
use.  Modern microprocessor chips contain multiple processing units
and multi-layer memories, for instance, and they use complicated
hardware/software strategies to move data and threads of computation
across those resources.  These features---along with all the others
that go into the design of these chips---make the patterns of their
processor loads and memory accesses highly complex and hard to
predict.  Accurate forecasts of these quantities, if one could
construct them, could be used to improve computer design.  If one
could predict that a particular computational thread would be bogged
down for the next 0.6 seconds waiting for data from main memory, for
instance, one could save power by putting that thread on hold for that
time period (e.g., by migrating it to a processing unit whose clock
speed is scaled back).  Computer performance traces are, however, very
complex.  Even a simple ``microkernel,'' like a three-line loop that
repeatedly initializes a matrix in column-major order, can produce
{\sl chaotic} performance traces \cite{mytkowicz09}, as shown in
Figure~\ref{fig:col-ts}, and chaos places fundamental limits on
predictability.
%
 \begin{figure}[htbp]
    \centering
    \includegraphics[width=\textwidth]{figs/colshortts}
    % where an .eps filename suffix will be assumed under latex,
    % and a .pdf suffix will be assumed for pdflatex
    \caption{A small snippet of the instructions per cycle(ipc) of {\tt
        \col}, a three-line C program that repeatedly initializes
      a matrix in column-major order, running on an Intel i7\textsuperscript{\textregistered}-based machine.  Even this
      simple program exhibits chaotic performance dynamics.}
   \label{fig:ipc}
  \end{figure}

\begin{figure}[htbp]
  \centering
  \begin{subfigure}[t]{0.475\textwidth}
    \includegraphics[width=0.95\textwidth]{figs/colshortts}
    \caption{A short segment of the instructions executed per CPU clock cycle
    (IPC) during the execution of \col. Each point is the IPC during a 100,000
    cycle period. There is a great deal of periodicity in this time series.}
    \label{fig:col-ts}
  \end{subfigure}%
  \begin{subfigure}[t]{0.475\textwidth}
    \includegraphics[width=0.95\textwidth]{figs/gccfullts}
    \caption{The full IPC time series during the execution of \gcc. This time
    series has very little structure.}
    \label{fig:gcc-ts}
  \end{subfigure}
  \label{fig:sample-ts}
\end{figure}

\begin{figure}[htbp]
  \centering

  \begin{subfigure}{0.5\textwidth}
    \includegraphics[width=\textwidth]{figs/gccARIMAForecast}
    \caption{\gcc ARIMA }
    \label{fig:gccARIMA}
  \end{subfigure}%
    \begin{subfigure}{0.5\textwidth}
    \includegraphics[width=\textwidth]{figs/gccLMAForecast}
    \caption{\gcc LMA}
    \label{fig:gccLMA}
  \end{subfigure}
  \\
  \begin{subfigure}{0.49\textwidth}
    \includegraphics[width=\textwidth]{figs/colARIMAForecast}
    \caption{\col ARIMA}
    \label{fig:colARIMA}
  \end{subfigure}
    \begin{subfigure}{0.49\textwidth}
    \includegraphics[width=\textwidth]{figs/colLMAForecast}
    \caption{\col LMA}
    \label{fig:colLMA}
  \end{subfigure}
  %\begin{subfigure}{0.5\textwidth}
  %  \includegraphics[width=1.0\textwidth]{figs/LMA_vs_ARIMA}
   \caption{[[move this to after the 3 forecast have been described and add the same plot for mean prediction]] 
For each of these, we plot the predicted value $p_n$ against the correct value $c_n$. On this type of plot a perfect prediction lies exactly on the diagonal, that is the line $p_n = c_n$, e.g., \ref{fig:colLMA} is a near perfect prediction where-as \ref{fig:gccARIMA} is a very poor prediction. }\label{fig:gcc_vs_col}  
  %\end{subfigure}
\end{figure}

The computer systems community has applied a variety of prediction
strategies to traces like this, most of which employ regression.  An
appealing alternative builds on the recently established fact that
computers can be effectively modeled as deterministic nonlinear
dynamical systems \cite{mytkowicz09}.  This result implies the
existence of a deterministic forecast rule for those dynamics.  In
particular, one can use \emph{delay-coordinate embedding} to
reconstruct the underlying dynamics of computer performance, then use
the resulting model to forecast the future values of computer
performance metrics such as memory or processor loads
\cite{josh-ida2011}.  In the case of simple microkernels like the one
that produced the trace in Figure~\ref{fig:ipc}, this deterministic
modeling and forecast strategy works very well.  In more-complicated
programs, however, such as speech recognition software or compilers,
this forecast strategy---as well as the traditional methods---break
down quickly.

This paper is a first step in understanding when, why, and how
deterministic forecast strategies fail when they are applied to
deterministic systems.  We focus here on the specific example of
computer performance.  We conjecture that the complexity of traces
from these systems---which results from the inherent dimension,
non-linearity, and non-stationarity of the dynamics, as well as from
measurement issues like noise, aggregation, and finite data
length---can make those deterministic signals \emph{effectively}
unpredictable.  We argue that \emph{permutation entropy}
\cite{bandt2002per}, a method for measuring the entropy of a
real-valued-finite-length time series through ordinal analysis, is an
effective way to explore that conjecture.  We study four
examples---two simple microkernels and two complex programs from the
SPEC benchmark suite---running on different Intel-based machines.  For
each program, we calculate the permutation entropy of the processor
load (instructions per cycle) and memory-use efficiency (cache-miss
rates), then compare that to the prediction accuracy attainable for
that trace using a simple deterministic model.

% paragraph to appease the theoretician in me
It is worth taking a moment to consider the theoretical possibility of
this task. We are not attempting to predict the state of the CPU at an
arbitrary point in the future --- this, at least with perfect
accuracy, would be tantamount to solving the halting problem. What we
are attempting is to predict aspects or functions of the running of
the CPU: instructions executed per second, cache misses per 100,000
instructions, and similar statistics. Prediction of these quantities
at some finite time in the future, even with perfect accuracy, does
not violate the Rice-Shapiro theorem.

% The rest of the paper is organized as follows.
% Section~\ref{sec:compModel} describes the experimental setup, as well
% as the nonlinear modeling and forecast strategies.  In
% Section~\ref{sec:meaComplex}, we review permutation entropy, calculate
% its value for a number of different computer performance traces, and
% compare the results to the prediction accuracy.  In
% Section~\ref{sec:conc}, we discuss these results and their
% implications in regard to our conjecture, and consider future areas of
% research.

\section{Experimental Methods }

This should be HPM PAPI, which programs we model

Include traces of full \col and \gcc

\section{MODELING}
\label{sec:compModel}

\subsection{LMA}
\subsection{ARIMA}
\subsection{Naive: Random Walk and mean}
\subsection{Prediction Accuracy}
MASE
Section outline:

\begin{enumerate}
\item Experimental methods (how we collect the time series and what the times series are
\item Description of DCE and parameter estimation  
\item Description of auto ARIMA 
\subitem (this should be limited and explain it is meant to be out of the box) point at the paper for auto arima for more details.
\item Description of the two naive methods (random walk and mean), make sure to explain that these methods are naive and simple but not necessarily bad.
\item Add a section talking about evaluation methods i.e., MASE, this text is currently written and just sitting at the beginning of the results.

\end{enumerate}

% took out for space
 \subsection{Reconstructing hidden dynamics}

Delay-coordinate embedding allows one to reconstruct a system's full
state-space dynamics from a \emph{single} scalar time-series
measurement---provided that some conditions hold regarding that data.
Specifically, if the underlying dynamics and the measurement
function---the mapping from the unknown state vector $\vec{X}$ to the
scalar value $x$ that one is measuring---are both smooth and generic,
Takens~\cite{takens} formally proves that the delay-coordinate map
\[
F(\tau,m)(x) = ([x(t) ~ x(t+\tau) ~ \dots ~x(t+m\tau)])
\]
from a $d$-dimensional smooth compact manifold $M$ to ${Re}^{2d+1}$,
where $t$ is time, is a diffeomorphism on $M$---in other words, that
the reconstructed dynamics and the true (hidden) dynamics have the
same topology.

This is an extremely powerful result: among other things, it means
that one can build a formal model of the full system dynamics without
measuring (or even knowing) every one of its state variables.  This is
the foundation of the modeling approach that is used in this paper.
The first step in the process is to estimate values for the two free
parameters in the delay-coordinate map: the delay $\tau$ and the
dimension $m$.  We follow standard procedures for this, choosing the
first minimum in the average mutual information as an estimate of
$\tau$ \cite{fraser-swinney} and using the false-near(est) neighbor
method of \cite{KBA92}, with a threshold of 10\%, to estimate $m$.  A
plot of the data from Figure~\ref{fig:ipc}, embedded following this
procedure, is shown in Figure~\ref{fig:embedding}.

 \begin{figure}
   \centering
     \includegraphics[width=\textwidth]{figs/colipc3d.png}
     \caption{A 3D projection of a delay-coordinate embedding of the trace
 from Figure~\ref{fig:ipc} with a delay ($\tau$) of 100,000 instructions.
 }
 \label{fig:embedding}
 \end{figure}

%% can cut for space if need be:
The coordinates of each point on this plot are differently delayed
elements of the \verb|col_major| L2 cache miss rate time series
$y(t)$: that is, $y(t)$ on the first axis, $y(t+\tau)$ on the second,
$y(t+2\tau)$ on the third, and so on.
%% ...down to here.
Structure in these kinds of plots---clearly visible in
Figure~\ref{fig:embedding}---is an indication of
determinism\footnote{A deeper analysis of
  Figure~\ref{fig:embedding}---as alluded to on the previous
  page---supports that diagnosis, confirming the presence of a chaotic
  attractor in these cache-miss dynamics, with largest Lyapunov
  exponent $\lambda_1 = 8000 \pm 200$ instructions, embedded in a
  12-dimensional reconstruction space \cite{mytkowicz09}.}.  That
structure can also be used to build a forecast model.

% took out for space
 \subsection{LMA: Using dynamics in forecasting}

Given a nonlinear model of a deterministic dynamical system in the
form of a delay-coordinate embedding like Figure~\ref{fig:embedding},
one can build deterministic forecast algorithms by capturing and
exploiting the geometry of the embedding.  Many techniques have been
developed by the dynamical systems community for this purpose
(e.g.,~\cite{casdagli-eubank92,weigend-book}).  Perhaps the most straightforward
is the ``Lorenz method of analogues'' (LMA), which is essentially
nearest-neighbor prediction in the embedded state
space~\cite{lorenz-analogues}.  Even this simple algorithm---which
builds predictions by finding the nearest neighbor in the embedded
space of the given point, then taking that neighbor's path as the
forecast---works quite well on the trace in Figure~\ref{fig:ipc}, as
shown in Figure~\ref{fig:cachePredTS}.
%
\begin{figure}[htbp]
  \centering
    \includegraphics[width=0.5\textwidth]{figs/colCachePredTS}
    \caption{A forecast of the last 4,000 points of the signal in
      Figure~\ref{fig:ipc} using an LMA-based strategy on the
      embedding in Figure~\ref{fig:embedding}.  Red circles and blue
      $\times$s are the true and predicted values, respectively;
      vertical bars show where these values differ. }
\label{fig:cachePredTS}
\end{figure}
%
On the other hand, if we use the same approach to forecast the
processor load\footnote{Instructions per cycle, or IPC} of the {\tt
  482.sphinx3} program from the SPEC cpu2006 benchmark suite, running
on an Intel i7\textsuperscript{\textregistered}-based machine, the
prediction is far less accurate; see Figure~\ref{fig:predsphinx}.
%
\begin{figure}[htbp]
  \centering
    \includegraphics[width=0.5\textwidth]{figs/sphinxPredicTS}
     \caption{An LMA-based forecast of the last 4,000 points of a
       processor-load performance trace from the {\tt 482.sphinx3}
       benchmark.  Red circles and blue $\times$s are the true and
       predicted values, respectively; vertical bars show where these
       values differ.}
\label{fig:predsphinx}
\end{figure}

Table~\ref{tab:PredError} presents detailed results about the
prediction accuracy of this algorithm on four different examples: the
{\tt col\_major} and {\tt 482.sphinx3} programs in
Figures~\ref{fig:cachePredTS} and~\ref{fig:predsphinx}, as well as
another simple microkernel that initializes the same matrix as {\tt
  col\_major}, but in row-major order, and another complex program
({\tt 403.gcc}) from the SPEC cpu2006 benchmark suite.  Both
microkernels were run on the Intel Core
Duo\textsuperscript{\textregistered} machine; both SPEC benchmarks
were run on the Intel i7\textsuperscript{\textregistered} machine.  We
calculated a figure of merit for each prediction as follows.  We held
back the last $k$ elements\footnote{Several different prediction
  horizons were analyzed in our experiment; the results reported in
  this paper are for $k$=4000} of the $N$ points in each measured time
series, built the forecast model by embedding the first $N-k$ points,
used that embedding and the LMA method to predict the next $k$ points,
then computed the Root Mean Squared Error (RMSE) between the true and
predicted signals:
$$RMSE = \sqrt{\frac{\sum_{i=1}^k(c_i-\hat{p_i})^2}{k}}$$
%
To compare the success of predictions across signals with different
units, we normalized RMSE as follows:
$$nRMSE = \frac{RMSE}{X_{max,obs}-X_{min,obs}}$$
%
% The smaller the nRMSE, obviously, the more accurate the prediction.

The results in Table~\ref{tab:PredError} show a clear distinction
between the two microkernels, whose future behavior can be
predicted effectively using this simple deterministic modeling
strategy, and the more-complex SPEC benchmarks, for which this
prediction strategy does not work nearly as well.
% Removed for space
% For both processor load (IPC) and memory usage (cache-miss rate),
% forecasts for {\tt 482.sphinx3} and {\tt 403.gcc} are much worse than
% for \verb|col_major| or \verb|row_major|.
%
This begs the question: If these traces all come from deterministic
systems---computers---then why are they not equally predictable?  Our
conjecture is that the sheer complexity of the dynamics of the SPEC
benchmarks running on the Intel i7\textsuperscript{\textregistered}
machine make them effectively impossible to predict.

%% [[If space: Add a segue sentence: next section uses permutation
%% entropy to explore that conjecture.]]

\subsection{Comparing Prediction Accuracy}
In order to analyze correctness of each prediction we split each time series into two pieces: the first 90\% referred to as the ``learning" or ``training" signal, $\{X_{i,obs}\}_{i=1}^{n}$ and the last 10\% known as the ``test" or ``correct" signal $\{c_j\}_{j=n+1}^{k+n+1}$. The learning signal is used to train an initial model (e.g., LMA or ARIMA) as described in Section \ref{sec:compModel}. The test signal is used both to assess the models forecasting accuracy and for any refitting that may be necessary. In particular, we perform $k$ 1-step predictions, after each 1-step prediction\footnote{We would like to note that this rebuilding occurs due to a problem with ARIMA models converging to a mean prediction if too long of a prediction horizon is used, this is not a handicap of either LMA or na\"ive.} we append the training signal with the next point in the correct signal $c_j$, refit the model taking into account the new system measurement and perform another prediction. This is repeated $k$ times to obtain $\{p_j\}_{j=n+1}^{k+n+1}$.

As a figure of merit we calculate the Mean Absolute Squared Error (MASE)\cite{MASE} between the true and predicted signals: 
%In order to compare the resulting forecasts we calculate the Mean Absolute Squared Error (MASE)\cite{MASE} between the true and predicted signals:
$$MASE = \sum_{j=n+1}^{k+n+1}\frac{|c_j-p_j| }{\frac{k}{n-1}\sum^n_{i=2}|X_{i,obs}-X_{i-1,obs}|}$$
The scaling term for MASE:
$$\frac{1}{n-1}\sum^n_{i=2}|X_{i,obs}-X_{i-1,obs}|$$ 
is the average in-sample forecast error for a random walk prediction $(p_i=X_{i-1,obs})$. This error method was introduced in \cite{MASE} as a ``generally applicable measurement of forecast accuracy without the problems seen in the other measurements." The major advantage of MASE is that it allows fair comparison across methods, prediction horizons and varying signal scales. When a forecast results in a $MASE<1$ this means that the prediction method gave, on average, smaller errors than the 1-step errors from the in-sample random walk forecast strategy. Analogously, $MASE>1$ means that the prediction method did worse, on average than the 1-step errors for the in-sample random walk forecast strategy. In Table \ref{tab:error} we provide the distribution [[Joshua: Ryan, Is this the right word? we give mean $\pm$ std. dev but some have very skewed right tails]]  of MASEs for each of the 8 signals and 3 prediction strategies, these are averaged over 15 runs of each type (signal + method). For comparison Table \ref{tab:error} also has the distribution of weighted permutation entropies for word lengths of $l=5$ and $l=6$.

\begin{table}[htdp]
\caption{[[Move this table to results, and cherry pick a few examples of \gcc and \col to put in the text and chop of $l=5$]] MASE distributions for 1-step predictions at a 10\% prediction horizon over 15 runs for each signal and average wpe at word length 5 and 6 for each signal.[[Joshua:Maybe delete $l=5$ as we don't use it in any figures, also if we leave this (before the next section we need to point forward to the next section for PE or remove it from this table]] }
\begin{center}
\begin{tabular}{|c|c|c|c|c|c|}
\hline
                   & MASE LMA    & MASE ARIMA &MASE na\"{i}ve & $l=5$  & $l=6$ \\
\hline
\gcc                  & $ 1.5296\pm 0.0214$ & $1.8366 \pm0.0157 $ & $1.7970\pm0.0095$& $0.9510 \pm 0.0011$ & $0.9430 \pm 0.0013$ \\

\col           & $ 0.0500 \pm0.0018  $ & $0.5989  \pm 0.2114 $ & $0.5707\pm0.0017$& $0.5636 \pm 0.0031$ & $0.5131 \pm 0.0034$ \\

\svd Reg. 1     & $ 0.8273\pm 0.0755$ & $ 0.7141\pm 0.0745 $ & $2.6763\pm4.3282$& $0.9761 \pm 0.0084$ & $0.9572 \pm 0.0156$ \\
\svd Reg. 2     & $1.2789 \pm0.0196 $ & $2.1626 \pm0.0265 $ &  $3.0543\pm0.0404$ &  $0.8760 \pm 0.0052$ & $0.8464 \pm0.0044$ \\
\svd Reg. 3       & $0.6192 \pm0.0209 $ & $0.7129 \pm 0.0096 $ & $31.3857\pm 0.2820$ & $0.7768 \pm 0.0073$ & $0.7157 \pm 0.0056$ \\
\svd Reg. 4     & $ 0.7789\pm0.0358 $ & $0.9787 \pm0.0321 $ & $2.6613\pm0.0739$                          &$0.9073 \pm 0.0080$ & $0.8246 \pm 0.0077$ \\
\svd Reg. 5     & $ 0.7177\pm 0.0483 $ & $2.3700  \pm 0.0505 $ & $20.8703 \pm 0.1915$& $0.7333 \pm 0.0076$ & $0.6776 \pm 0.0068$ \\
\svd Reg. 6     & $ 0.7393\pm 0.0682 $ & $ 1.4379\pm 0.0609$ & $2.1967\pm0.0830$& $0.8101 \pm 0.0135$ & $0.7475 \pm 0.0106$ \\
\hline
\end{tabular}
\end{center}
\label{tab:error}
\end{table}%

\section{Measuring Structural Complexity}\label{sec:meaComplex}

THINGS TO DO
\begin{enumerate}
\item rewrite for consistency in notation. For example, some parts of the paper call word length $l$ and some call it $n$.

\item make sure it is clear and we justify that detecting and characterizing structural complexity is hard for real-valued noisy time series.
\item Quantifying structured and unstructured complexity is nontrivial in the case of real-valued noisy time series but WPE does this. [[talk about again here but justify in information theory section]]
\item justify and explain the regime split in \svd 
\end{enumerate}

For the purposes of this paper, one can view entropy as a measure of complexity
and predictability in a time series.  A high-entropy time series is almost
completely unpredictable---and conversely.  This can be made more rigorous:
Pesin's relation \cite{pesin77} states that in chaotic dynamical systems, the
Shannon entropy rate is equal to the sum of the positive Lyapunov exponents,
$\lambda_i$. The Lyapunov exponents directly quantify the rate at which nearby
states of the system will diverge with time: $\left| \Delta x(t) \right| \approx
e^{\lambda t} \left| \Delta x(0) \right|$.  The faster the divergence, the more
difficult prediction becomes.

Utilizing entropy as a measure of temporal complexity is by no means a new idea
\cite{Shannon1951, mantegna1994linguistic}.  Its effective usage requires
categorical data: $x_t \in \mathcal{S}$ for some finite or countably infinite
\emph{alphabet} $\mathcal{S}$, whereas data taken from real-world systems is
effectively real-valued.  To get around this, one must discretize the data---
typically achieved by binning.  Unfortunately, this is rarely a good solution to
the problem, as the binning of the values introduces an additional dynamic on
top of the intrinsic dynamics whose entropy is desired.  The field of symbolic
dynamics studies how to discretize a time series in such a way that the
intrinsic behavior is not perverted, but these methods are fragile in the face
of noise and require further understanding of the underlying system, which
defeats the purpose of measuring the entropy in the first place.

Bandt and Pompe introduced the \emph{permutation entropy} (PE) as a ``natural
complexity measure for time series" \cite{bandt2002per}.  Permutation entropy
employs a method of discretizing real-valued time series that follows the
intrinsic behavior of the system under examination.  Rather than looking at the
statistics of sequences of values, as is done when computing the Shannon
entropy, permutation entropy looks at the statistics of the \emph{orderings} of
sequences of values using ordinal analysis. Ordinal analysis of a time series is
the process of mapping successive time-ordered elements of a time series to
their value-ordered permutation of the same size.  By way of example, if $(x_1,
x_2, x_3) = (9, 1, 7)$ then its \emph{ordinal pattern}, $\phi(x_1, x_2, x_3)$,
is $231$ since $x_2 \leq x_3 \leq x_1$.  This method has many features; among
other things, it is generally robust to observational noise and requires no
knowledge of the underlying mechanisms.

\begin{mydef}[Permutation Entropy]

  Given a time series $\{x_t\}_{t = 1,\dots,T}$. Define $\mathcal{S}_n$ as all
  $n!$ permutations $\pi$ of order $n$. For each $\pi \in \mathcal{S}_n$ we
  determine the relative frequency of that permutation occurring in $\{x_t\}_{t
  = 1,\dots,T}$:
  \begin{align*}
    p(\pi) = \frac{\left|\{t|t \leq T-n,\phi(x_{t+1},\dots,x_{t+n}) = \pi\}\right|}{T-n+1}
  \end{align*}
  Where $|\cdot|$ is set cardinality. The \emph{permutation entropy} of order $n
  \ge 2$ is defined as
  \begin{align*}
  H(n) = - \sum_{\pi \in \mathcal{S}_n} p(\pi) \log_2 p(\pi)
  \end{align*}

\end{mydef}

Notice that $0\le H(n) \le \log_2(n!)$ \cite{bandt2002per}.  With this in mind,
it is common in the literature to normalize permutation entropy as follows:
$\frac{H(n)}{\log_2(n!)}$.  With this convention, ``low'' entropy is close to 0
and ``high'' entropy is close to 1. Finally, it should be noted that the
permutation entropy has been shown to be identical to the Shannon entropy for
many large classes of systems \cite{amigo2012permutation}.

Here we will be utilizing a variation of the permutation entropy, the
\emph{weighted permutation entropy} (WPE)~\cite{fadlallah2013}. The weighted
permutation entropy attempts to correct for observational noise which is larger
than some trends in the data, but smaller than the larger scale features --- for
example, a signal that switches between two fixed points with noise about those
fixed points. The weighted permutation entropy would be dominated by the
switching rather than by the stochastic fluctuation. To accomplish this, the
\emph{weight} of a permutation is taken into account:
\begin{align*}
  w(x_{t+1:t+n}) = \frac{1}{n}
                 \sum_{x_i \in \{x_{t+1:t+n}\}}
                 \left( x_i - \bar{x}_{t+1:t+n} \right)^2
\end{align*}
where $x_{t+1:t+n}$ is a sequence of values $x_{t+1}, \ldots, x_{t+n}$, and
$\bar{x}_{t+1:t+n}$ is the arithmetic mean of those values.

The weighted probability of a permutation is then:
\begin{align*}
  p_w(\pi) = \frac{\displaystyle \sum_{t \le T - n} w(x_{t+1:t+n}) \cdot \delta(\phi(x_{t:t+n}), \pi) }{\displaystyle \sum_{t \le T - n} w(x_{t+1:t+n})}
\end{align*}
where $\delta(x, y)$ is 1 if $x = y$ and 0 otherwise. Effectively, this weighted
probability enhances permutations involved in ``large'' features and demotes
permutations which are small in amplitude relative to the features of the time
series. The weighted permutation entropy is then:
\begin{align*}
  H_w(n) = - \sum_{\pi \in \mathcal{S}_n} p_w(\pi) \log_2 p_w(\pi),
\end{align*}
which can also be normalized by dividing by $\log_2(n!)$, and will be in all the
results of this paper.

In practice, calculating permutation entropy and weighted permutation entropy
involves choosing a good value for the word length $n$. The primary
consideration is that the value be large enough that forbidden ordinals are
discovered, yet small enough that reasonable statistics over the ordinals are
gathered: e.g.:
\begin{align*}
  n = \operatornamewithlimits{argmax}_\ell \{ T \gtrapprox 100 \ell! \},
\end{align*}
assuming an average of 100 counts per ordinal is sufficient. In the literature,
$3 \le n \le 6$ is a standard choice --- generally without any formal
justification. In theory, the permutation entropy should reach an asymptote with
increasing $n$, but that requires an arbitrarily long time series. In practice,
what one should do is calculate the \emph{persistent} permutation entropy by
increasing $n$ until the result converges, but data length issues can intrude
before that convergence is reached.

The weighted permutation entropy for the {\tt SVD} program is given in
Fig.~\ref{fig:wwpe}. To generate this image a window of 5,000 values slid over
the time series. Within each of those windows, the statistics over words of
length 4 are computed and the WPE is calculated. The gray bands denote regions
where the 5,000 value window overlapped visually-distinct regimes. It can be
seen that the behaviors of the weighted permutation entropy vary between
regimes. [[I think here it would be good to add a paragraph explaining the windowed WPE was used for regime choices on SVD...emphasizing  that over a time series permutation entropy fluctuates illustrating within a single time series different levels of complexity and predictability exist. Maybe point at some of the predicting predictability papers.]]

%\begin{figure}[htbp]
%  \centering
%  \includegraphics[width=1.0\textwidth]{figs/SVD_wwpe}
%  \caption{[[Joshua: I think adding the colored SVD trace to this would be %good or putting it above this figure]]The weighted permutation entropy of %one run of SVD. The gray bands
%    are regions where the window overlaps regimes. The window size used is
%    $5,000 \times 100,000$ instructions and the word length is $4$.}
%  \label{fig:wwpe}
%\end{figure}

%%FIGURE INTENT
%This figure shows that entropy changs over time with the signal as well as why the regimes for \svd were chosen. 
%%%%%%%%%%%%%%%%%
\begin{figure}[htbp]
  \centering

  %\begin{subfigure}{0.3\textwidth}
  %  \includegraphics[width=0.3\textwidth]{figs/svdipcregimescolored.png}
  %  \caption{The instructions per cycle of \svd. Each color corresonds to the different regimes as selected by rapid shifts in WPE, as seen in Figure \ref{fig:svd_wwpe}. From left to right each change in color represents a change in regime for 6 regimes in total. }
 %   \label{fig:svd_ts}
 % \end{subfigure}%
  %\\
 % \begin{subfigure}{\textwidth}
    \includegraphics[width=\textwidth]{figs/SVD_wwpe}
    %\caption{The MASE of ARIMA vs weighted permutation entropy. }
  \caption{
[Joshua: I think adding the colored SVD trace to this would be good or putting it above this figure but need to figure how to line them up properly. Also we need to label that the numbers on the bottom of WPE are regimes not instructions...]]The weighted permutation entropy of one run of SVD. The gray bands
    are regions where the window overlaps regimes. The window size used is
    5,000 $\times$ 100,000 instructions and the word length is $4$.[[Joshua:why is $l=4$ and not 5,6 like the rest of the paper?]]For reference the instructions per cycle of \svd are plotted as a ghost behind this plot. Each color on the ghosted time series corresonds to the different regimes as selected by rapid shifts in WPE. From left to right each change in color represents a change in regime for 6 regimes in total.}\label{fig:wwpe}
\end{figure}

 \section{Validation of Primary Findings} % (fold)
 \label{sec:results}

%OUTLINE:
%\begin{enumerate}
%\item \cmark~Introduce MASE.
%\item \cmark~WPE is a good measure of predictability.  %Figure: best athlete MASE vs.
%its WPE.
%\item \cmark~Talk about structure analysis. Just because %there is forward information
%transfer, does not mean that linear predictors can get at %this.
%\subitem \cmark~For this show a figure of (a) ARIMA vs MASE %(b) LMA vs MASE in a side by%
%side plot.
%\item full results.  Image: MASE vs. WPE for both LMA \& %ARIMA.  Points to make:
%\begin{enumerate}
%\item \cmark~clusters are distributed differently
%\item \cmark~clusters are shaped differently---tight or not
%\item \cmark~clusters move differently between LMA and ARIMA
%\item finally, the diagonal line is important. If you're %below it, you could do
%better.
%\end{enumerate}
%\end{enumerate}

%BRAINSTORMING 
%\begin{itemize}
%\cmark\item The kind of complexity present matters, i.e., that is whether the complexity is structured or not.[[use here and mention in intro]] 
%\cmark\item Quantifying structured and unstructured complexity is nontrivial in the case of real-valued noisy time series but WPE does this. [[talk about again here but justify in information theory section]]
%\item Maybe plot a big chunk of \col and a big chunk of \gcc together and show that they both look complex.

%\cmark\item \gcc appears visually very complex, *and* according to WPE this complexity is unstructured. And a constant, linear and nonlinear prediction strategy all fail. We should be able to conclude that guessing random values is the best we can do as is shown by MASE [[Use in this section]]

%\cmark\item \col is also complex (can even be chaotic/ point to CHAOS paper) but the complexity is structured according to WPE and as such that complexity is usable for prediction [[use in this section]]

%\cmark\item \col brings about the point nicely that some prediction strategies cannot utilize the processes internal information transfer method. That is a nonlinear internal information transfer system cannot be predicted effectively with a linear strategy. This gives a practitioner leverage on when to give up and when to keep working. [[use in this section as bridge to next section]]

%\end{itemize}

In this section we validate the two key primary findings of this work which we introduced in Section \ref{sec:intro}:

\begin{enumerate}

\item The existence of predictable structure in noisy real-valued time series is quantifiable by WPE and as a result WPE is correlated with prediction accuracy (MASE) of an ideal predictor.

%\subitem [[Simply reminders not to be included]]WPE can quantify when a noisy real-valued time series is predictable.[[I am unsatisfied with predictable in this sentence, need a better word to say "better than random walk" or ``able to be forecast effectively" or ``has the structural capacity to transmit information in a way that the time series can be effectively forecast"

%\item The existence of complex structure in a noisy real-valued time series is quantifiable and this type of complexity is directly correlated with predictability

\item The way structure/information/complexity is processed internally by a given process plays a crucial role in predictability.

%\subitem [[Simply reminders not to be included]]We will have shown that the existence whether linear or nonlinear is picked up on with WPE but this point gets at whether the prediction model can use the structure or not (linear can't use nonlinear structure). The right to left shifts in \col and some of the \svd regimes and the lack of shift in \gcc illustrate this nicely.

%\subitem [[Simply reminders not to be included]]This is the linear vs nonliear vs random we see with the right to left shifts with lower complexity time series.

\end{enumerate}

%This portion should justify the following claim%%%%%%%%%%%%%%%%%%%%%%
%\item The existence of predictable structure in noisy real-valued time series is quantifiable by WPE and as a result WPE is correlated with prediction accuracy (MASE)  
%%%%%%%%%%%%%%%%%%%%%%%%%

%\item First paragraph: WPE is a good measure of predictability.  Figure:
%best athlete MASE vs. its WPE.

%\item Quantifying structured and unstructured complexity is nontrivial in the case of real-valued noisy time series but WPE does this. [[talk about again here but justify in information theory section]]

As we discussed in Section \ref{sec:meaComplex} distinguishing structured from unstructured complexity in the case of real-valued time series is non trivial, i.e., distinguishing when a complex signal omits predictable structure and when the signal is effectively random is not a trivial task. As described in Section \ref{sec:intro}, for a practitioner this can be frustrating because it can be nearly impossible to find the source of faulty predictions: Is it simply that we need to use a more ideal (possibly more advanced) predictor or is it the case that the time series is simply so complex that using a simple (yet inconsistent) forecast strategy such as random walk is the best we can do. Fortunately, weighted permutation entropy (WPE) allows us to make this distinction for noisy real-valued time series.

In Figure~\ref{fig:pred_vs_wpe} we plot the best prediction (i.e., the lowest MASE over all 3 methods\footnote{LMA,ARIMA,na\"ive} over all 15 runs of that program) for each of the 8 programs under investigation. 
\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.75\textwidth]{figs/prediction_vs_entropy}
  \caption{The best MASE among all runs and prediction methods vs weighted
  permutation entropy. For each of these, the word length used is $6$. The
  dashed line is a least-squares linear fit of all the points except for {\tt
  SVD$_1$} which we have excluded for reasons explained in the text.}
  \label{fig:pred_vs_wpe}
\end{figure}
What we see here is that the relationship between prediction accuracy (MASE) and the weighted permutation entropy is much as we conjectured: The existence of predictable structure in noisy real-valued time series as quantified by low to moderate WPE  is correlated with prediction accuracy (MASE) of the ``best"predictor\footnote{We do not claim to have found the ideal predictor for each signal, simply the best predictor over what we used; Which we believe to be a fair sampling of standard prediction techniques for this type of signal.}. We will not further validate this claim through some examples.

>>>>

\input{intro.tex}

\input{methods.tex}

\input{model.tex}

\input{wpe.tex}

\input{results.tex}

\input{concl.tex}
